A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-16 review/response.qmd

# Reviewer: 2

This paper proposes new pipeline work for index development, including index generation, validation, and sensitivity analysis. The authors develop a new R package, "tidyindex" including all functions in the pipeline to produce a new index. This package will be beneficial for developing new indices. The central part of this paper is the development of the R package and the explanation of its usage. However, this package needs to be fixed.

The authors explain their functions briefly in Chapter 5 and use examples with R code in Chapter 6. However, the code in Chapter 6 does not work. Also, the demo in the R package (CRAN version 0.1.0) did not work (see attached R-output.html). Please check the package, including demos and examples in this paper. It would be great if all R codes for the example in Chapter 6 were provided, including generating plots.

**\textcolor{purple}{Fixed. We were expecting that readers worked from the vignettes folder. The problem was that the demo folder had some irrelevant code. We've removed these now. We have also re-organised the code for the paper as follows.}**

- **\textcolor{purple}{The main file tidyindex.R, reproduces the content in the paper from saved data objects, available in the data folder.}** 
- **\textcolor{purple}{New scripts have been provided in the scripts folder which completely generate each of the examples. There is a README.md, that instructs the reader how to set their environment up to reproduce all of the analyses.}**
- **\textcolor{purple}{The GitHub repo is listed in the Acknowledgements section, and any further updates needed for future versions to run will be made there.}**

@@

## minor comments

- In Table 1, it would be great if there were a line between items. Because the name of the module is in two lines, it is not easy to match the other information (input, operation, and output)

**\textcolor{purple}{Fixed. The table layout has been modified accordingly.}**.

 - On page 19, " The output contains the original data, index values (.index), parameters used (.scale,
.method, and .dist), and all the intermediate variables (.pet, .agg, and .fitted)." However, the output has no .index, .pet, .agg, and .fitted. Please check.

**\textcolor{purple}{Fixed. The output has been written in a way that all variables are visible now (using dplyr::glimpse()).}**.

- Figure 7 shows the sequence of screenshots from the animation to explore the sensitivity of the GGGI. It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest",

**\textcolor{purple}{Frame 12 is described in more detail now, and additional explanation of the information being communicated by the figure is provided.}**

- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004. It would be great if the author mentioned it, too.

**\textcolor{purple}{Added. This wider confidence interval is interesting and corresponds to the millennium drought in Australia.}**

- reference for "SPEI package" on page 4 should be (Beguería and Vicente-Serrano 2017),
and reference for "the Standardized Precipitation-Evapotranspiration Index (SPEI)" on page 12 should be (Vicente-Serrano, Beguería, and López-Moreno 2010)

**\textcolor{purple}{Fixed}**.

### 2024-04-16 tidyindex.qmd
tidied intro and conclusions

<!-- - Define what is an index, what is not -->

@@

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. They serve as critical tools across the natural and social sciences. Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index. In environmental science, climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.  In economics, indexes provide insight into market trends through combining prices of a basket of goods and services. In social sciences, indexes are used to monitor human development, gender equity, or university quality. The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.

To construct an index, experts typically start by defining a concept of interest that requires measurement. This concept often lacks a direct measurable attribute or can only be measured as a composite of various processes, yet it holds social and public significance. To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims to measure the process of interest. The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others. For instance, the indexes constructed from a linear combination of variables require a decision on the weight assigned to each variable. Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes. Hence, all these decisions can result in different index values and have different practical implications.

To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component. We call this process of breaking the index construction into different steps the index pipeline. Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes,  calculating index uncertainty, and assessing index robustness.

@@

The paper introduces a novel data pipeline for constructing and analyzing indexes within the tidy framework. The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.

One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity. The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains. These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility. This work highlights the importance of carefully constructing statistical indexes and introduces novel computational methods for studying and calculating statistical indexes and their uncertainties.

One future development of the work is to extend the tidyindex framework to support index calculation on commonly used domain-specific data formats, for example, NetCDF for climate indexes.

@@

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship. Nicolas Langrené acknowledges the partial support of the Guangdong Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up Research Fund UICR0700041-22. The article is created using Quarto [@Allaire_Quarto_2022] in R [@R]. The source code for reproducing this paper can be found at: <https://github.com/huizezhang-sherry/paper-tidyindex>.

>>>>

@@

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. They serve as critical tools across the natural and social sciences. Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index. In environmental science, climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.  In economics, indexes provide insight into market trends through combining prices of a basket of goods and services. In social sciences, indexes are used to monitor human development, gender equity, or university quality.

<!-- what is the gap - why is this statistics -->
The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared. This violates *tidy principles* [@tidyverse] which makes it difficult to investigate the nature of calculations through a principled data pipeline. It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@???Spiegelhalter, someone else].

<!-- - Define what is an index, what is not -->
To construct an index, experts typically start by defining a concept of interest that requires measurement. This concept often lacks a direct measurable attribute or can only be measured as a composite of various processes, yet it holds social and public significance. To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims to measure the process of interest. The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others. For instance, the indexes constructed from a linear combination of variables require a decision on the weight assigned to each variable. Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes. Hence, all these decisions can result in different index values and have different practical implications.

To be able to test different decision choices for an index, systematically and statistically, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component. We call this process the *index pipeline*, which are the steps of the data analysis pipeline for the index construction. Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness.

@@

The paper introduces a tidy data pipeline for constructing and analyzing indexes. It has nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.

This work addresses aspects often overlooked in index construction: uncertainty quantification and sensitivity. The pipeline supports confidence interval calculation and sensitivity analysis to be incorporated in the index construction process. It should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.

The pipeline approach is illustrated with examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility. This work highlights the importance of carefully constructing statistical indexes and introduces new computational methods for studying and calculating statistical indexes and their uncertainties.

There are many potential directions for the development of the work. Computationally, the tidyindex framework could be extended to support other commonly used data formats, for example, NetCDF for climate indexes. Conceptually, extending the examples to re-express additional common-practice indexes in the pipeline structure will help broader adoption, and further test that the framework can indeed accommodate any and all possible indexes.

@@

The steps of this pipeline are available in the R package, `tidyindex`, available on CRAN. The source code for reproducing the work reported in this paper are in Supplementary Materials and can be found at: <https://github.com/huizezhang-sherry/paper-tidyindex>.

This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship. Nicolas Langrené acknowledges the partial support of the Guangdong Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up Research Fund UICR0700041-22. The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
